---
title: "Data Cleaning"
author: "Geovanny Risco"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  github_document:
    keep_html: no
    number_sections: yes
    toc: yes
    toc_depth: 3
    df_print: tibble
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, rows.print = 25, fig.path = "figures/")
```

# Import libraries

```{r}
library(tidyverse)
```

# Constant/Config variables

```{r}
batch_number <- "_batch3"
MAX_NULLS_PER_ANTIBIOTIC <- 30 # In percentage
```

# Import data

```{r}
# ARGs (Antibiotic Resistance Genes)
args_data_filepath <- paste0("data/results/resfinder/args_table", batch_number, ".csv")
args_data <- read_csv(args_data_filepath, col_types = cols("sample_name" = col_character()))
## Fix/extract name of samples
args_data <- args_data %>%
  mutate(sample_name = str_extract(sample_name, "^\\w+.\\d+"))
args_data
```

```{r}
# Results from CARD database
card_data_filepath <- paste0("data/results/card/card_results", batch_number, ".tsv")
card_data <- read_tsv(card_data_filepath, na = c("n/a"))
# Fix/extract name of samples
card_data <- card_data %>%
  mutate(SAMPLE_ID = str_extract(SAMPLE_ID, "^\\w+.\\d+"))
card_data
```

```{r}
# AMR (Antimicrobial Resistance) labels
amr_labels <- read_csv("data/results/data_collection_ncbi/amr_labels.csv", col_types = cols("SampleID" = col_character()))
amr_labels
```

```{r}
# Load NCBI samples metadata
samples_metadata <- read_tsv("data/results/data_collection_ncbi/assembly_accession_ids+tax_ids.txt", col_names = c("biosample_accession", "assembly_accession", "tax_id"))
samples_metadata
```

For now and until BV-BRC is active again, we will filter out 1351
samples (which are not in NCBI). In order to add BV-BRC samples, we need
to parse AMR labels information in a different way:

```{r}
amr_labels <- amr_labels %>%
  filter(`SampleID` %in% samples_metadata$biosample_accession)
card_data <- card_data %>%
  filter(SAMPLE_ID %in% samples_metadata$assembly_accession)
args_data <- args_data %>%
  filter(sample_name %in% samples_metadata$assembly_accession)
```

# Clean and prepare data

First of all, we will need to clean and prepare the data in order to
perform the analysis.

## ARGs from Resfinder

This table has the following structure:

| sample_name     | GeneA | GeneB | GeneC | ... |
|:----------------|-------|-------|-------|----:|
| GCA_012637185.1 | 0     | 1     | 0     | ... |
| ...             | ...   | ...   | ...   | ... |

For each gene, a boolean value is given dependending on whether the gene
is resistance or not.

-   1: resistance gene
-   0: non-resistance gene

### Overview of the data
  
```{r}
# Sum of all different ARG genes (sum all columns that have at least one 1)
args_data %>%
  summarise_all(~ sum(.x == 1, na.rm = TRUE)) %>%
  gather(key = "gene", value = "count") %>%
  filter(count > 0) %>%
  arrange(desc(count))
```

### Null values detection

```{r}
# Count number of nulls per sample
args_data %>%
  mutate(nulls = rowSums(is.na(select(., -sample_name)))) %>%
  select(sample_name, nulls) %>%
  arrange(desc(nulls))
```

We have no nulls values for ARGs.

### Outliers detection

```{r}
# For each sample, how many resistance genes are present?
args_data %>%
  mutate(count = rowSums(select(., -sample_name))) %>%
  select(sample_name, count) %>%
  arrange(desc(count))
# Summary metrics for number of resistance genes per sample
args_data %>%
  mutate(count = rowSums(select(., -sample_name))) %>%
  summarise(mean = mean(count), std = sd(count), median = median(count), min = min(count), max = max(count))
```

```{r "Boxplot for distribution of ARGs per sample"}
# Boxplot summarizing above information
args_data %>%
  mutate(count = rowSums(select(., -sample_name))) %>%
  ggplot(aes(x = "", y = count)) +
  geom_boxplot() +
  labs(x = "", y = "Number of resistance genes") +
  theme_classic()
n_args_samples <- nrow(args_data)
```

As can be seen in the boxplot, there are some samples that have a strange high number of resistance genes.
These outliers can produce noise in the data, so we will need to set a theshold to remove them.

```{r}
# Set threshold to remove outliers
args_outliers_threshold <- 15
```

We will consider as anormal samples those that have more than `r args_outliers_threshold` resistance genes.
We don't know the reason behind this abnormality though, we will need to investigate further.

```{r "Boxplot for distribution of ARGs per sample (without outliers)"}
# Remove those samples that have more than the threshold of resistance genes
args_data <- args_data %>%
  mutate(count = rowSums(select(., -sample_name))) %>%
  filter(count <= args_outliers_threshold) %>%
  select(-count)
n_args_samples_without_outliers <- nrow(args_data)

# Remake boxplot
args_data %>%
  mutate(count = rowSums(select(., -sample_name))) %>%
  ggplot(aes(x = "", y = count)) +
  geom_boxplot() +
  labs(x = "", y = "Number of resistance genes") +
  theme_classic()
```

After applying the threshold, we have removed a total of `r n_args_samples - n_args_samples_without_outliers` samples.

```{r "Histogram with distribution of ARGs per sample"}
# Histogram with density of resistance genes per sample
args_data %>%
  mutate(count = rowSums(select(., -sample_name))) %>%
  ggplot(aes(x = count)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "#000000", fill = "#0099F8") +
  geom_density(color = "#000000", fill = "#F85700", alpha = 0.6) +
  labs(x = "Number of resistance genes", y = "Density") +
  theme_classic()
```

Recalculate summary metrics after applying filters:

```{r}
args_data %>%
  mutate(count = rowSums(select(., -sample_name))) %>%
  summarise(mean = mean(count), std = sd(count), median = median(count), min = min(count), max = max(count))
```

### Feature engineering

As all of the columns we have in this table is boolean data, there is
not much to do in terms of feature engineering. Hoewever, we have
noticed that there are some ARGs that does not contribute to resistance
in any sample, that is, they are always 0. Since this does not provide
us any information, we will remove them from the table.

```{r}
# Check length of args_data columns
original_ncols <- length(colnames(args_data))

# Filter to only columns that has any other value different than 0
args_data <- args_data %>%
  select_if(function(x) any(x != 0))

removed_ncols <- original_ncols - length(colnames(args_data))
```

After the filtering, we have removed `r removed_ncols` columns.

## SNPs from CARD

Although [CARD database](https://card.mcmaster.ca/) offers us a large
variety of information about AMR vectors, we will only use the SNPs
information. For more information about the output format, please refer
to the official [documentation](https://github.com/arpcard/rgi#id72).

### Filtering

We will be filtering by the following criteria: 

* Column `Model_type`
must be either `protein variant model` or `protein overexpression model`

* They must have a value within the column `SNPs_in_Best_Hit_ARO`.
NOTE: this column can have multiple values separated by commas.

```{r}
# Filter by Model_type
card_snps_data <- card_data %>%
  filter(Model_type %in% c("protein variant model", "protein overexpression model"))

# Filter by SNPs_in_Best_Hit_ARO
card_snps_data <- card_snps_data %>%
  filter(!is.na(SNPs_in_Best_Hit_ARO))

# Explode SNPs_in_Best_Hit_ARO
card_snps_data <- card_snps_data %>%
  mutate(SNPs_in_Best_Hit_ARO = strsplit(SNPs_in_Best_Hit_ARO, ",")) %>%
  unnest(SNPs_in_Best_Hit_ARO)
```

### Exploration/Visualization

```{r summary_data}
# Summary metrics number of SNPs per sample
card_snps_data %>%
  group_by(SAMPLE_ID) %>%
  summarise(n_snps = n()) %>%
  summarise(mean = mean(n_snps), std = sd(n_snps), median = median(n_snps), min = min(n_snps), max = max(n_snps))
```

```{r "Boxplot for distribution of SNPs per sample"}
# Boxplot showing how many SNPs are present in each sample
card_snps_data %>%
  group_by(SAMPLE_ID) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = "", y = count)) +
  geom_boxplot() +
  labs(x = "", y = "Number of SNPs") +
  theme_classic()
```

Recalulate summary metrics after applying threshold:

```{r}
card_snps_data %>%
  group_by(SAMPLE_ID) %>%
  summarise(n_snps = n()) %>%
  summarise(mean = mean(n_snps), std = sd(n_snps), median = median(n_snps), min = min(n_snps), max = max(n_snps))
```

```{r "Histogram of distribution of SNPs per sample"}
# Histogram showing how many SNPs are present in each sample
card_snps_data %>%
  group_by(SAMPLE_ID) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = count)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "#000000", fill = "#0099F8") +
  geom_density(color = "#000000", fill = "#F85700", alpha = 0.6) +
  labs(x = "Number of SNPs", y = "Density") +
  theme_classic()
```

### Preparation

Now that we have filtered the data, we will need to transform it into a
format compatible for ML algorithms, that is, a table with the features
of interest as columns and the samples as rows. In this case, the
features we are interested in are the SNPs, so we will need to pivot the
table so that each row represents a sample and each column represents a
SNP ID (column `SNPs_in_Best_Hit_ARO`). The value of each cell will be
the number of times that the SNP appears in the sample.

```{r}
# Pivot table
card_snps_data_wide <- card_snps_data %>%
  select(SAMPLE_ID, SNPs_in_Best_Hit_ARO) %>%
  group_by(SAMPLE_ID, SNPs_in_Best_Hit_ARO) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = SNPs_in_Best_Hit_ARO, values_from = count, values_fill = 0) %>%
  ungroup()
```

## AMR labels

The structure of this table is as follows:

| SampleID     | Antibiotic1 | Antibiotic2 | Antibiotic3 | ... |
|:-------------|-------------|-------------|-------------|----:|
| SAMN04256112 | 0           | 1           | 0           | ... |
| ...          | ...         | ...         | ...         | ... |

Where each column represents an antibiotic and each row represents a
sample. The values of each cell can be:

-   0: the sample is not resistant to the antibiotic
-   1: the sample is resistant to the antibiotic

One sample can be resistant to multiple antibiotics, so we can have
multiple 1s in the same row.

### Preparation

Adapt data so it has the same sampleIds as ARGS and variant calling
data. AMR labes happens to have the biosamples accession numbers as
sampleIds, so we will need to map them to their corresponding assembly
accession ids.

```{r}
# Given information in samples_metadata, replace biosample_accession with assembly_accession
amr_labels <- amr_labels %>%
  left_join(distinct(samples_metadata, biosample_accession, .keep_all = TRUE), by = c("SampleID" = "biosample_accession")) %>%
  select(-c(`SampleID`, tax_id)) %>%
  rename(`SampleID` = assembly_accession) %>%
  select(SampleID, everything())
```

### Cleaning

#### Null values analysis

We will remove those antibiotics with more than
`r MAX_NULLS_PER_ANTIBIOTIC`% of null values.

```{r}
# Count null values per antibiotic (each column) in percentage
nulls_per_antibiotic <- amr_labels %>%
  select(-`SampleID`) %>%
  summarise_all(~ sum(is.na(.x)) / nrow(amr_labels) * 100) %>%
  gather(key = "antibiotic", value = "% of null values") %>%
  arrange(desc(`% of null values`))
knitr::kable(nulls_per_antibiotic)
```

```{r}
# Remove antibiotics with more than 30% of null values
antibiotics_to_remove <- nulls_per_antibiotic %>%
  filter(`% of null values` > MAX_NULLS_PER_ANTIBIOTIC) %>%
  pull(antibiotic)
amr_labels <- amr_labels %>%
  select(-all_of(antibiotics_to_remove))

# Fill remaining null values with 0
amr_labels <- amr_labels %>%
  mutate_all(~ replace_na(.x, 0))
```

A total of `r length(antibiotics_to_remove)` antibiotics have been removed.

#### Outliers analysis

```{r}
# Count number of resistant antibiotics per sample
amr_labels %>%
  mutate(n_resistant = rowSums(select(., -`SampleID`))) %>%
  select(`SampleID`, n_resistant) %>%
  arrange(desc(n_resistant))
```

```{r "Boxplot for distribution of resistant antibiotics per sample"}
# Boxplot summarizing above information
amr_labels %>%
  mutate(n_resistant = rowSums(select(., -`SampleID`))) %>%
  ggplot(aes(x = "", y = n_resistant)) +
  geom_boxplot() +
  labs(x = "", y = "Number of antibiotics") +
  theme_classic()
```

```{r "Histogram with distribution of resistant antibiotics per sample"}  
# Histogram with density of resistant antibiotics per sample
amr_labels %>%
  mutate(n_resistant = rowSums(select(., -`SampleID`))) %>%
  ggplot(aes(x = n_resistant)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "#000000", fill = "#0099F8") +
  geom_density(color = "#000000", fill = "#F85700", alpha = 0.6) +
  labs(x = "Number of antibiotics", y = "Density") +
  theme_classic()
```

We do not observe relevant outliers in this case, so we will not apply any threshold.

### Exploration

Count how many samples are resistance to each antibiotic:

```{r, display = 'inline-block'}
resistant_samples_per_antibiotic <- amr_labels %>%
  select(-`SampleID`) %>%
  summarise_all(~ sum(.x == 1, na.rm = TRUE)) %>%
  gather(key = "antibiotic", value = "resistant samples") %>%
  arrange(desc(`resistant samples`))
knitr::kable(resistant_samples_per_antibiotic)
```

```{r, echo=FALSE}
# Customize colour palette
antibiotics_color_palette <- c("#8DD3C7", "#FFFFB3", "#BEBADA", "#FB8072", "#80B1D3", "#FDB462", "#B3DE69", "#FCCDE5", "#D9D9D9", "#BC80BD", "#CCEBC5", "#FFED6F", "lightblue")
```

```{r n_resistant_samples_per_antibiotic}
# Bar plot with number of resistant samples per antibiotic
resistant_samples_per_antibiotic %>%
  ggplot(aes(x = antibiotic, y = `resistant samples`, fill = antibiotic)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  labs(x = "Antibiotic", y = "Number of resistant samples") +
  scale_fill_manual(values = antibiotics_color_palette) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0, 4000))
```

# Explore data

Median number of resistant genes per antibiotic:

```{r median_number_resistant_genes_per_antibiotic}
# Boxplot with median number of resistant genes per antibiotic
args_data %>%
  mutate(n_args = rowSums(select(., -sample_name))) %>%
  select(sample_name, n_args) %>%
  left_join(amr_labels, by = c("sample_name" = "SampleID")) %>%
  pivot_longer(cols = -c(sample_name, n_args), names_to = "antibiotic", values_to = "resistant") %>%
  filter(resistant == 1) %>%
  ggplot(aes(x = antibiotic, y = n_args, fill = antibiotic), ) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  labs(x = "Antibiotic", y = "Number of resistant genes") +
  scale_fill_manual(values = antibiotics_color_palette) +
  scale_y_continuous(limits = c(0, 20)) +
  theme(legend.position = "none")
```

Median number of CARD SNPs per antibiotic:

```{r median_number_card_snps_per_antibiotic}
# Boxplot with median number of CARD SNPs per antibiotic
card_snps_data %>%
  group_by(SAMPLE_ID) %>%
  summarise(n_card_snps = n()) %>%
  left_join(amr_labels, by = c("SAMPLE_ID" = "SampleID")) %>%
  pivot_longer(cols = -c(SAMPLE_ID, n_card_snps), names_to = "antibiotic", values_to = "resistant") %>%
  filter(resistant == 1) %>%
  ggplot(aes(x = antibiotic, y = n_card_snps, fill = antibiotic)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  labs(x = "Antibiotic", y = "Number of SNPs") +
  scale_fill_manual(values = antibiotics_color_palette) +
  scale_y_continuous(limits = c(0, 10)) +
  theme(legend.position = "none")
```

# Save data

```{r}
card_snps_data_output_path <- paste0("data/results/card/card_snps_data", batch_number, "_cleaned.tsv")
card_snps_data_latest_output_path <- "data/results/card/card_snps_data_latest_cleaned.tsv"
card_snps_data_wide %>%
  write_tsv(card_snps_data_output_path)
card_snps_data_wide %>%
  write_tsv(card_snps_data_latest_output_path)

args_data_output_path <- paste0("data/results/resfinder/args_data", batch_number, "_cleaned.tsv")
args_data_latest_output_path <- "data/results/resfinder/args_data_latest_cleaned.tsv"
args_data %>%
  write_tsv(args_data_output_path)
args_data %>%
  write_tsv(args_data_latest_output_path)

amr_labels_output_path <- paste0("data/results/data_collection_ncbi/amr_labels", batch_number, "_cleaned.tsv")
amr_labels_latest_output_path <- "data/results/data_collection_ncbi/amr_labels_latest_cleaned.tsv"
amr_labels %>%
  write_tsv(amr_labels_output_path)
amr_labels %>%
  write_tsv(amr_labels_latest_output_path)
```
